---
title: "Targeted MSDial Pipeline: QE"
author: "RLionheart"
date: "12/26/2019"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: inline
---

This markdown script controls the targeted pipeline for QE Cyano and HILIC data run through MSDial metabolomics software.

It contains four major sections:
Section I: Import and cleaning/rearranging of data.
Section II: Quality control using user-defined parameters.
Section III: Applying Best-Matched Internal Standard (B-MIS).
Section IV: Quantifying peak area to umol/vial when possible.

TODO (rlionheart):
- Change import sections in files to flexible function.


```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(anytime)
library(rlist)
library(tidyverse)
library(tidyr)
options(scipen=999)

source("src/Functions.R")

processed.folder <- "data_processed" 
figures.folder <- "figures"
intermediate.folder <- "data_intermediate"
dir.create(file.path(".", processed.folder))  
dir.create(file.path(".", figures.folder))  
dir.create(file.path(".", intermediate.folder))  

if (length(dir(path = "data_processed/")) == 0) {
  cat("\nData_processed subdirectory has been created and is ready for new data.")
} else {
  cat("Warning: data_processed subdirectory is not empty. Empty contents before continuing.\n")
}
```


If you would like to empty the data_processed/, data_intermediate, and figures/ subdirectories, run the following code.
```{r, include = TRUE}
toClear <- c("data_processed/", "data_intermediate", "figures/")
f <- list.files(toClear, include.dirs = F, full.names = T, recursive = T)
file.remove(f)

print("Subdirectories emptied.")
```

----------------------------------------------------------------------------------------------------------------------------

Section I: Import all MSDial files that have been split by Area, Mass/charge (Mz), Retention Time (RT), and Signal to Noise (SN).

In the MSDial datasets:
Set header, filter unknowns.
Change variable classes from character/factor to numeric, according to column.
Rearrange the dataframes and combine to a single frame in long format.
Standardize dataset by removing "Ingalls_" prefixes from compound names, and removing the syntactically correct "X" from Replicates.

***
Inputs: 
"data_raw/*file.pattern*.csv
Outputs: 
"data_intermediates/MSDial_combined_*file.pattern*_*DATE*.csv"
***

*User action required*
Comment or uncomment the file.pattern and matching.pattern required for your files.

```{r Pattern matching, include = TRUE}
# Cyano
# file.pattern <- "CYANO"
# matching.pattern <- "RP.Cyano"

# HILIC
file.pattern <- "HILIC"
matching.pattern <- "positive|negative"

print(paste(file.pattern, "has been selected."))
```

```{r MSDial Imports, include = TRUE}
source("src/MSDial_Import.R")

print("Required files imported.")
```


*User action required*
Enter the existing filenames of your run. The above code assigns the variables in R to their filename in the directory, so if your positive Area file is "data_processed/PositiveArea.csv", it will be imported to this code as PositiveArea. Those files need to be reassigned to Area.positive so the rest of the code will know which files to edit for the pipeline steps.

Comment or uncomment the block of variable names appropriate for your run.

```{r Dataset reassignment, include = TRUE}
# Comment out the run not being used.

# Cyano variables: 
# Area.RP.Cyano <- Area_CYANO_EddyTransect
# Mz.RP.Cyano   <- Mz_CYANO_EddyTransect
# RT.RP.Cyano   <- RT_CYANO_EddyTransect
# SN.RP.Cyano   <- SN_CYANO_EddyTransect

# HILIC variables: 
Area.positive <- Area_HILICPos_Example
Mz.positive   <- Mz_HILICPos_Example
RT.positive   <- RT_HILICPos_Example
SN.positive   <- SN_HILICPos_Example

Area.negative <- Area_HILICNeg_Example
Mz.negative   <- Mz_HILICNeg_Example
RT.negative   <- RT_HILICNeg_Example
SN.negative   <- SN_HILICNeg_Example

print(paste(file.pattern, "variables assigned."))

```

Rearrange and export.
```{r Dataset rearrangement, include = TRUE}
source("src/MSDial_Rearrange.R")

print("Data rearrange complete.")
```

--------------------------------------------------------------

Section II: Quality Control and flagging of problematic peaks.

In the Quality Control Step:
Import files.
Identify run types and check if all are present (blk, smp, std, poo).
Create a table of standard retention times (RT) for comparison.
Create a table of areas from blank runs for comparison.
Flag peaks in the dataset that fall outside of user-defined bounds.
Add parameter values to the top of the final file and save to the data_processed/ folder.

***
Inputs: 
"data_intermediate/MSDial_combined_*file.pattern*_*DATE*.csv"

Outputs: 
"data_intermediate/MSDial_RT.table_*DATE*.csv"
"data_intermediate/MSDial_final.table_*DATE*.csv"
"data_intermediate/MSDial_blank.table_*DATE*.csv"
"data_processed/MSDial_QC_Output_*file.pattern*_*DATE*.csv"
***

*User action required*
Define parameters for quality control. These act as comparison for filtering out data.
The numbers will change depending on whether you are analyzing HILIC or Cyano data.
```{r QC parameters, include = TRUE}
area.min   <- 1000
RT.flex    <- 0.4
blk.thresh <- 0.3
SN.min     <- 4

print("Parameter values assigned.")
```

Run Quality Control and export.
```{r MSDial QC, include=TRUE}
source("src/MSDial_QC.R")
```

Inspect the blank.table, final.table, and RT.table values, which currently exist in the environment.
Ensure that they look normal before proceeding to clear the environment in the next step.
```{r, include = TRUE}
currentDate <- Sys.Date()
csvFileName <- paste("data_processed/MSDial_QC_Output_", file.pattern, "_", currentDate, ".csv", sep = "")

tables <- grep("table", names(.GlobalEnv), value = TRUE, ignore.case = TRUE)
tablelist <- do.call("list", mget(tables))

# Write intermediate data
invisible(lapply(tables, 
                 function(x) write.csv(get(x), file=paste("data_intermediate/MSDial_", x, "_", currentDate, ".csv", sep = ""))))
# Write final data
write.csv(final.table, csvFileName, row.names = FALSE)

print(paste(tables, "saved to data/intermediate"))

rm(list = setdiff(ls()[!ls() %in% c("file.pattern")], lsf.str()))
```

--------------------------------------------------------------

Section III: Best-Matched Internal Standard (B-MIS)

In the BMIS step:
If working with HILIC data, identify and remove duplicates. Justify your decision.
Match QC'd data with Internal Standards list.
Identify internal standards and visualize those areas.
Test if replicate names are identical across the data and the sample key. Stop the analysis process if not.
Identify/visualize internal standards. Look over it to ensure that things are looking more or less "normal".
Using pooled and sample runs, calculate internal standard averages. 
Create an adjusted area for all pooled and sample runs by comparing each mass feature to each internal standard. 
Calculate RSD for each mass feature/internal standard pair, and use this to choose a matched internal standard.
Decide whether to accept a B-MIS or not, according to user-defined values.

***
Inputs:
"data_extras/*Sample_Key_from_instrument*.csv"
"data_extras/Ingalls_Lab_Standards.csv"
"data_processed/MSDial_QC_Output_*file.pattern*_*DATE*.csv"

Outputs:
"figures/IS.Raw.Areas.png"
"figures/BMIS_Evalution.png"
"data_intermediate/MSDial_InternalStdIssues_*DATE*.csv"
"data_intermediate/MSDial_QuickReport_*file.pattern*_*DATE*.txt"
"data_processed/MSDial_BMIS_Output_*file.pattern*_*DATE*.csv"
***

*User action required*
Enter user data for cut off parameters.
cut.off = Decrease in RSD of pooled injections, aka improvement cutoff.
cut.off2 = Relative squared deviation minimum.

```{r BMIS cutoff values, include = TRUE}
cut.off <- 0.4 
cut.off2 <- 0.1 

print("B-MIS Cutoff values assigned.")
```

*User action required*
Comment out appropriate variable blocks according to HILIC or Cyano data.
```{r Cyano HILIC assignment, include = TRUE}
# Cyano
# Column.Type = "RP"
# standards.pattern = "Ingalls"
# QC.pattern = "QC_Output_CYANO"

# HILIC
Column.Type = "HILIC"
sample.key.pattern = "HILIC"
standards.pattern = "Ingalls"
QC.pattern = "QC_Output_HILIC"

print(paste(file.pattern, "column type, standards pattern, and QC pattern assigned."))
```

Import required files.
```{r BMIS imports, include=TRUE}
source("src/MSDial_BMIS_Imports.R")

print("BMIS files imported.")
```

If data is HILIC, identify duplicates and decide which to remove 
IdentifyDuplicates function will confirm if instrument column data exists.
User will need to use best judgement to decide which duplicate to remove.
```{r Check duplicates, include=TRUE}
source("src/MSDial_Duplicates.R")
currentDate = Sys.Date()
csvFileName <- paste("data_intermediate/MSDial_HILIC.duplicates_", currentDate, ".csv", sep = "")

# Using the duplicates.testing table, decide which detected compound to keep, the positive or negative.

if (exists("duplicates.testing") == TRUE) {
  print("This is a HILIC run. Look at duplicates.testing to decide which compounds to remove.")
} else {
  print("Non-HILIC data: no duplicates to remove.")
}

```

*User action required*
The below section automatically removes the HILICNeg compounds from the QC'd data. 
This is not necessarily the correct solution: the user must look at the data to ensure that they are removing the right compounds.
```{r Remove duplicates, include = TRUE}
if (exists("HILICS.duplicates")) {
  QCd.data <- QCd.data %>%
    filter(!(Metabolite.Name %in% HILICS.duplicates$Metabolite.Name & Column == "HILICNeg")) # This line requires user attention.
  print("HILICS duplicates removed.")

  Report <- print(paste(HILICS.duplicates$Metabolite.Name, "HILICNeg"))
  cat(Report, file = "data_intermediate/MSDIAL_HILIC_DuplicateReport.txt")
  
  write.csv(duplicates.testing, csvFileName, row.names = FALSE)
} else {
  print("Non-HILIC data: no duplicates removed.")
}

```

Run BMIS.
```{r BMIS, include=TRUE}
source("src/MSDial_BMIS.R")
```

--------------------------------------------------------------

Section IV: Convert from peak area to umol/vial.

In the quantify step:
Get response factors and response factor ratios.
Quantify samples without an internal standard.
Quantify samples with an internal standard.
Accounting for dilution and filtered volume, calculate environmental quantities.
Summarize carbon and nitrogen.

***
Inputs:
"data_extras/Ingalls_Lab_Standards.csv"
"data_extras/InternalStandardNames.csv"
"data_processed/BMIS_Output_*file.pattern*_*DATE*.csv"
"data_processed/MSDial_QC_Output_*file.pattern*_*DATE*.csv"

Outputs:
"data_intermediate/MSDial_ResponseFactorRatios_*DATE*.csv"
"data_processed/Quantified_Summary_*Column.Type*_*DATE*.csv
"data_processed/Quantified_Measurements*Column.Type*_*DATE*.csv
"data_processed/Quantified_perSampID_*Column.Type*_*DATE*.csv
***

*User action required*
Enter dilution factor, injection volume, and the filtered volume from the instrument run.
```{r, include = TRUE}
Dilution.Factor = 2
Injection.Volume = 400 # nanomoles
Volume.Filtered = 5 # liters

print("Dilution factor, injection volume, and filtered volume assigned.")
```

*User action required*
Comment out appropriate variable blocks according to HILIC or Cyano data.
```{r, include = TRUE}
# Cyano
# standards.pattern = "Ingalls"
# BMIS.pattern = "BMIS_Output_RP"
# QC.pattern = "QC_Output_CYANO"
# names.pattern = "Names"
# Column.Type = "RP"

# HILIC
standards.pattern = "Ingalls"
BMIS.pattern = "BMIS_Output_HILIC"
QC.pattern = "QC_Output_HILIC"
names.pattern = "Names"
Column.Type = "HILIC"

print(paste(file.pattern, "matching patterns assigned."))

```

Import required files for quantification.
```{r Quantify imports, include = TRUE}
source("src/MSDial_Quantify_Imports.R")

print("Quantification files imported.")
```

Repeat the HILIC duplicates step.
```{r Check duplicates, include=TRUE}
check.duplicates <- list.files("data_intermediate", pattern = "duplicate", full.names = TRUE)

if (length(check.duplicates) == 0) {
  print("No HILIC data exists.")
} else {
  print("HILIC data exists.")
  
  HILICS.duplicates <- read.csv(check.duplicates, stringsAsFactors = FALSE)
}
```


*User action required*
If you are working with HILIC data, make sure your filtering step makes sense.
```{r Check duplicates, include=TRUE}

if (length(check.duplicates) == 0) {
  print("No HILIC data exists.")
} else {
  HILICS.duplicates <- read.csv(check.duplicates, stringsAsFactors = FALSE)
  QCd.data <- QCd.data %>%
    filter(!(Metabolite.Name %in% HILICS.duplicates$Metabolite.Name & Column == "HILICNeg"))
  print("Duplicates removed.")
}

```

Check which kinds of standards have been run.
*****************************************************************************
This function is unlikely to work on all runs due to sampID differences.
It will almost definitely need to be modified to apply to the dataset at hand.
*****************************************************************************
```{r, include = TRUE}
Full.data <- CheckStandards2(Full.data)
```

Quantify.
```{r, include = TRUE}
source("src/MSDial_Quantify.R")
```

Review the "Final" files to ensure everything is correct.

Save and export files.
```{r, include = FALSE}
currentDate <- Sys.Date()
csvFileName.summed <- paste("data_processed/Quantified_Summary_", Column.Type, "_", currentDate, ".csv", sep = "")
csvFileName.final <- paste("data_processed/Quantified_Measurements_", Column.Type, "_", currentDate, ".csv", sep = "")
csvFileName.perID <- paste("data_processed/Quantified_perSampID_", Column.Type, "_", currentDate, ".csv", sep = "")


write.csv(Final.Quantitative.Summed, csvFileName.summed, row.names = FALSE)
write.csv(Final.Quantitative, csvFileName.final, row.names = FALSE)
write.csv(Final.All.perSampID, csvFileName.perID, row.names = FALSE)
```

Clear environment.
```{r, include=FALSE}
rm(list = ls())
```